---
title: "Aggregation example"
output: html_notebook
---

November, 2020

Replication code for empirical example in the Appendix of
"How do populations aggregate?" by Dennis M. Feehan and Elizabeth Wrigley-Field

NOTE: in order to run this code, you will have to download the life tables from the
[United States Mortality Database](https://usa.mortality.org/).
This requires that you register (which is free). 

The file you need is the 
[bundled archive called lifetables.zip](https://usa.mortality.org/uploads/lifetables/lifetables.zip).

This analysis is based on the version of `lifetables.zip` that was downloaded from the USMD website on
March 21, 2019.

```{r}
library(tidyverse)
library(HMDHFDplus)
library(patchwork)
library(here)
```

Create directories for data and output

```{r}
dir.create(here("raw-data"), showWarnings=FALSE)
dir.create(here("out"), showWarnings=FALSE)

# this should be whichever directory has the raw USMD data
usmd.data.dir <- here("raw-data")

# this is where we'll save outputs
out.dir <- here("out")
```

This chunk will ask you for your US Mortality Database username and password.
Then it will download the lifetables for you.

```{r}
usmd_uname <- readline("Your username:")
usmd_pwd <- readline("Your password:")

httr::GET(url = "https://usa.mortality.org/uploads/lifetables/lifetables.zip",
          httr::authenticate(password=usmd_pwd, user=usmd_uname),
          httr::write_disk(file.path(usmd.data.dir, 'lifetables.zip')))

```

Load the data

```{r}
zipped_csv_names <- grep('\\.csv$', unzip(file.path(usmd.data.dir, 'lifetables.zip'), list=TRUE)$Name, 
                           ignore.case=TRUE, value=TRUE)

state_codes <- c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", 
"GA", "HI", "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", 
"ME", "MI", "MN", "MO", "MS", "MT", "NC", "ND", "NE", "NH", "NJ", 
"NM", "NV", "NY", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", 
"TX", "UT", "VA", "VT", "WA", "WI", "WV", "WY")
```

```{r}
# quick function to get the life table for a state
get_lt <- function(code, type='b', fmt="1x1", zipfn = file.path('raw-data', 'lifetables.zip')) {
  fullpath <- glue::glue("lifetables/States/{code}/{code}_{type}ltper_{fmt}.csv")
  this_lt <- read_csv(unz(zipfn, fullpath), 
                      col_types='ccdcdddddddd') %>%
             mutate(Age = HMDHFDplus::age2int(Age))
  return(this_lt)
}

# read all the lifetables in
lts_both <- setNames(map(state_codes, ~get_lt(.x, 'b')), state_codes)
lts_male <- setNames(map(state_codes, ~get_lt(.x, 'm')), state_codes)
lts_female <- setNames(map(state_codes, ~get_lt(.x, 'f')), state_codes)

lts <- bind_rows(bind_rows(lts_both), bind_rows(lts_male), bind_rows(lts_female))

# save this to avoid having to re-load the data all the time
# (this is not necessary for replication, so commenting it out)
#write_csv(lts, file=file.path(out.dir, 'usmd_lts.csv'))
```


```{r}
# weighted harmonic mean of x, weights given by w
hmean <- function(x, w) {
  w.o.x <- w/x
  return(sum(w) / sum(w.o.x))
}
```

In these life tables, mx is not exactly equal to dx / Lx. We'll add a column that is.

```{r}
lts <- lts %>% mutate(dx.over.Lx = dx/Lx)
```

Go through and calculate reldiff by age and sex when aggregation is incorrect

```{r}
reldiffs <- lts %>%
  filter(Year == 2015) %>%
  split(list(.$Sex, .$Age)) %>% 
  map_df(function(asdat) {
    
    ## only look at states with at least 1 death
    asdat <- asdat %>% filter(dx > 0)
    
    direct.agg.mx <- sum(asdat$dx) / sum(asdat$Lx)
    
    # harmonic mean of rates, weighted by deaths (correct)
    hm.mx.d <- hmean(asdat$dx.over.Lx, asdat$dx)
    # arithmetic mean of rates, weighted by nLx (correct)
    am.mx.L <- weighted.mean(asdat$dx.over.Lx, asdat$Lx)
    
    # arithmetic mean of rates, weighted by deaths (incorrect)
    am.mx <- weighted.mean(asdat$dx.over.Lx, asdat$dx)
    # arithmetic mean of rates, unweighted (incorrect)
    uam.mx <- mean(asdat$dx.over.Lx)
    
    # arithmetic mean of death rates, weighted by exposure, squared
    am.m.d2 <- weighted.mean(asdat$dx.over.Lx, asdat$Lx)^2
    # arithmetic mean of squared death rates, weighted by exposure
    am.m2.d <- weighted.mean(asdat$dx.over.Lx^2, asdat$Lx)
    
    # appropriately weighted cv2.mx
    w.cv2.mx <- (am.m2.d/am.m.d2) - 1
    
    # the weighted cv (not squared)
    w.cv.mx <- sqrt(w.cv2.mx)
    
    # diff in arithmetic mean weighted by deaths, relative to true aggregate value
    reldiff <- (am.mx - direct.agg.mx) / direct.agg.mx
    
    # diff in unweighted arithmetic mean, relative to true aggregate value
    reldiff2 <- (uam.mx - direct.agg.mx) / direct.agg.mx
    
    return(tibble(age=asdat$Age[1],
                  sex=asdat$Sex[1],
                  n=nrow(asdat),
                  direct=direct.agg.mx,
                  # correct harmonic mean
                  hm=hm.mx.d,
                  # arithmetic mean with wrong weights
                  am=am.mx,
                  # arithmetic mean with no weights
                  uam=uam.mx,
                  reldiff.am = reldiff,
                  reldiff.uam = reldiff2,
                  w.cv.mx = w.cv.mx,
                  w.cv2.mx = w.cv2.mx))
  }) %>%
  mutate(sex = dplyr::recode(sex,
                             'b'='Both sexes',
                             'm'='Males',
                             'f'='Females')) 
```


```{r}
fig.h <- 5
fig.w <- 8

am.reldiffs <- ggplot(reldiffs) +
  geom_point(aes(x=age, y=reldiff.am, color=sex)) +
  expand_limits(y=c(0,.6)) +
  theme_minimal() +
  ylab("Relative error") +
  ggtitle("Relative error\nusing arithmetic mean\nweighted by deaths")
#ggsave(plot=am.reldiffs, filename='am-reldiffs.pdf', height=fig.h, width=fig.w)
am.reldiffs
```

```{r}
uam.reldiffs <- ggplot(reldiffs) +
  geom_point(aes(x=age, y=reldiff.uam, color=sex)) +
  expand_limits(y=c(0,.6)) +
  theme_minimal() +
  ylab("Relative error") +
  ggtitle("Relative error\nusing arithmetic mean\nunweighted")
#ggsave(plot=uam.reldiffs, filename='uam-reldiffs.pdf', height=fig.h, width=fig.w)
uam.reldiffs
```

```{r}
fig.h <- 5
fig.w <- 8

relerr.plot <- (uam.reldiffs | am.reldiffs) + plot_annotation(tag_levels = 'A')
ggsave(plot=relerr.plot, filename=file.path(out.dir, 'relerrs.pdf'), height=fig.h, width=fig.w)
```

Plot the actual aggregate death rates

```{r}
agg.pseudous <- ggplot(reldiffs) +
  geom_point(aes(x=age, y=direct, color=sex)) +
  theme_minimal() +
  scale_y_log10() +
  ylab("nMx") +
  ggtitle("True aggregate\nage-specific death rate\n(logged)")
#ggsave(plot=agg.pseudous, filename='agg-pseudo-us.pdf', height=fig.h, width=fig.w)
agg.pseudous
```

Plot the variance in means

```{r}
cv.pseudous <- ggplot(reldiffs) +
  geom_point(aes(x=age, y=w.cv.mx, color=sex)) +
  theme_minimal() +
  ylab("CV") +
  ggtitle("Coefficient of variation in\ndeath rates\nacross sub-populations")

cv.pseudous
```


```{r}
fig.h <- 5
fig.w <- 8

pseudo.m.plot <- (agg.pseudous | cv.pseudous) + plot_annotation(tag_levels = 'A')
ggsave(plot=pseudo.m.plot, filename=file.path(out.dir, 'pseudous-m.pdf'), height=fig.h, width=fig.w)
```


Directly compare rel-sd and error in second strategy

```{r}
cv2.relerr.plot <- ggplot(reldiffs) +
  #geom_point(aes(x=w.cv2.mx, y=reldiff.am, color=log(direct))) +
  geom_abline(intercept=0, slope=1, color='grey') +
  geom_point(aes(x=w.cv2.mx, y=reldiff.am)) +
  #xlim(0,.1) + ylim(0,.1) +
  xlab("Squared coefficient of variation in death rates") +
  ylab("Relative error\nusing arithmetic mean\nweighted by deaths") +
  theme_minimal() +
  coord_equal()

cv2.relerr.plot
```
```{r}
fig.h <- 5
fig.w <- 5

ggsave(plot=cv2.relerr.plot, 
       filename=file.path(out.dir, 'cv2-relerr.pdf'), height=fig.h, width=fig.w)
```
